---
title: "NTL_LTER"
output: pdf_document
date: "2024-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(vegan)
library(ggpubr)
```


## Potential problems
(1) abundance cannot represent biomass; for phytoplankton it has no problems, but for upper trophic levels body size should be considered. 
(2) when calculating stability, do we need to convert date-based sampling events to annual average first, because each year may have different sampling events?
(3) I still need stability data for phytoplankton; I will use primary productivity data;


```{r phytoplankton in trout lake area, fig.width=4, fig.height=3}
library(tidyverse)
library(vegan)
rm(list=ls())
##
data_phyto_trout <- read.csv('harmonized_data/phytoplankton_trout_lake_area.csv')
##
## The Simpson index is a quantitative measure of how evenly individuals are distributed among species in a community.the problem with "- 1" if abundance is not count. 
##
## The Shannon index: H = -SUM(p * ln (p)). It measures the diversity of a community. 
##
## Steps of doing so:
## (0.0) check unit_abundance and scale_abundance are uniform throughout a dataset; otherwise uniform unit is needed. 
## (0.0) convert this excel sheet to large matrix
## (1) annual average of community characteristics
## (2) aggregated stability matrix
## (3) average of the whole data
## (4) diversity statistics (I should use these packages)
##
## Do we have duplicated rows? Yes. deal with duplication!
sum(duplicated(data_phyto_trout[,1:14])==TRUE)
##
## remove duplicated rows
data_phyto_trout <- data_phyto_trout %>% group_by(year, month, day, unique_ID, taxon_name) %>% summarise(abundance = sum(abundance))
##
## convert it into a data.frame
data_phyto_trout_convert <- data_phyto_trout %>% mutate(date=as.Date(year, month, day)) %>% pivot_wider(names_from=taxon_name, values_from = abundance, values_fill = 0)
## change NA values into 0
data_phyto_trout_convert[is.na(data_phyto_trout_convert)] <- 0.0
##
## when to use group
data_phyto_trout_annual <- data_phyto_trout_convert %>% ungroup %>% dplyr::select(-(c('date', 'month', 'day'))) %>% group_by(unique_ID, year) %>% summarise_all(mean)
##
## calculate stability, but this data has only one year data
##
## further step to site level data
data_phyto_trout_site <- data_phyto_trout_annual %>% ungroup %>% dplyr::select(-year) %>% group_by(unique_ID) %>% summarise_all(mean)
H <- diversity(data_phyto_trout_site[,-1])
S <- specnumber(data_phyto_trout_site[,-1])
## they are highly correlated. 
plot(H, S)
##
diversity_stability_phyto <- data.frame(unique_ID = data_phyto_trout_site$unique_ID, shannon = H, richness = S, lake_id=substring(data_phyto_trout_site$unique_ID, 1, 2))
#
lake_feature <- read.csv('lake_characteristics.csv')
diversity_stability_phyto <- left_join(diversity_stability_phyto, lake_feature, by="lake_id")
##
## diversity is quite similar among these sites. 
##
## get stability of primary productivity
productivity <- read.csv('/Users/junnawang/courses/SSECR/data_discovery/North Temperate Lake LTER/knb-lter-ntl.402.2_Metabolism Phenology/model_output.csv')
### We have the productivity for 5 sites ~ "allequash", "bigmuskellunge", "crystal", "trout", "sparkling"; I can calculate this. 
unique(productivity$lake)
###
stability_nep <- productivity %>% group_by(lake) %>% summarise(cv_1 = abs(mean(Fnep))/sd(Fnep))
stability_nep$lake_id <- c('AL', 'BM', 'CR', 'fish', 'mendota', 'monona', 'TR', 'SP')
### great! the five lakes has 38 years data
diversity_stability_phyto <- diversity_stability_phyto %>% left_join(stability_nep, by='lake_id')
diversity_stability_phyto$area_log <- log(diversity_stability_phyto$surface.area..hectare.)
#
p1 <- ggplot(data=diversity_stability_phyto, aes(x=shannon, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Producer shannon diversity', y='Producer aggregated stability (1/cv)', title='Producer (phytoplankton)')

p2 <- ggplot(data=diversity_stability_phyto, aes(x=richness, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Producer richness', y='Producer aggregated stability', title='Producer (phytoplankton)')

p3 <- ggplot(data=diversity_stability_phyto, aes(x=area_log, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Log transformed lake surface area', y='Producer aggregated stability (1/cv)', title='Producer (phytoplankton)')

ggarrange(p1, p2, p3, nrow=1, ncol=3)
ggsave("project_results/producer_diversity_stability.png")
#

ggplot(data=diversity_stability_phyto, aes(x=richness, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green', label.x = 45, label.y = 1.5, size = 5) +
  labs(x='Total richness of producers', y='Producer aggregated stability', title='Producer (phytoplankton)')


```

```{r zooplankton trout lake, fig.width=4, fig.height=3}
#
# This includes data in 1980s and 1990s. 
data_zoop_trout <- read.csv('harmonized_data/zooplankton_trout_lake_area.csv')
# unifying lake id
data_zoop_trout$unique_ID[data_zoop_trout$unique_ID=='Tr_lake_2'] <- 'TR_lake_2'
data_zoop_trout$unique_ID[data_zoop_trout$unique_ID=='Bm_lake_1'] <- 'BM_lake_1'
##
## remove the rows with species name = empty: 218 rows
data_zoop_trout <- data_zoop_trout[data_zoop_trout$taxon_name!="", ]
#
sum(duplicated(data_zoop_trout[,1:14])==TRUE)
## 15 duplicated rows
#
## remove duplicated rows
data_zoop_trout <- data_zoop_trout %>% group_by(year, month, day, unique_ID, taxon_name) %>% summarise(abundance = sum(abundance))
##
## convert it into a data.frame
data_zoop_trout_convert <- data_zoop_trout %>% mutate(date=as.Date(year, month, day)) %>% pivot_wider(names_from=taxon_name, values_from = abundance, values_fill = 0)
## change NA values into 0
data_zoop_trout_convert[is.na(data_zoop_trout_convert)] <- 0.0
##
## aggregated into annual average
data_zoop_trout_annual <- data_zoop_trout_convert %>% ungroup %>% dplyr::select(-(c('date', 'month', 'day'))) %>% group_by(unique_ID, year) %>% summarise_all(mean)
##
## calculate stability
data_zoop_trout_annual$total <- rowSums(data_zoop_trout_annual[, -(1:2)])
stability_cv <- data_zoop_trout_annual %>% group_by(unique_ID) %>% summarise(cv_1 = mean(total)/sd(total))
##
## further step to site level data
data_zoop_trout_site <- data_zoop_trout_annual %>% ungroup %>% dplyr::select(-year) %>% group_by(unique_ID) %>% summarise_all(mean)
H <- diversity(data_zoop_trout_site[,-1])
S <- specnumber(data_zoop_trout_site[,-1])
## they are highly correlated. 
plot(H, S)
##
diversity <- data.frame(unique_ID = data_zoop_trout_site$unique_ID, shannon = H, richness = S, lake_id=substring(data_zoop_trout_site$unique_ID, 1, 2))
##
## combine diversity and stability
diversity_stability_zoop <- diversity %>% left_join(stability_cv, by="unique_ID")
##
## adding the effect of ecosystem size
lake_feature <- read.csv('lake_characteristics.csv')
diversity_stability_zoop <- diversity_stability_zoop %>% left_join(lake_feature, by='lake_id')
plot(data=diversity_stability_zoop, log(surface.area..hectare.) ~ cv_1)    # stability is highly correlated to lake area
##
diversity_stability_zoop$area_log <- log(diversity_stability_zoop$surface.area..hectare.)
summary(lm(data=diversity_stability_zoop, cv_1 ~ area_log + shannon))
summary(lm(data=diversity_stability_zoop, cv_1 ~ shannon))
summary(lm(data=diversity_stability_zoop, cv_1 ~ richness))
summary(lm(data=diversity_stability_zoop, cv_1 ~ area_log))           # this is the best model. 
##
p1 <- ggplot(data=diversity_stability_zoop, aes(x=shannon, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Consumer shannon diversity', y='Consumer aggregated stability (1/cv)', title='Consumer (zooplankton)')

p2 <- ggplot(data=diversity_stability_zoop, aes(x=richness, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Consumer richness', y='Consumer aggregated stability (1/cv)', title='Consumer (zooplankton)')

p3 <- ggplot(data=diversity_stability_zoop, aes(x=area_log, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Log transformed lake surface area', y='Consumer aggregated stability (1/cv)', title='Consumer (zooplankton)')

ggarrange(p1, p2, p3, nrow=1, ncol=3)
ggsave("project_results/consumer_zooplankton_diversity_stability.png")

ggplot(data=diversity_stability_zoop, aes(x=richness, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green', label.x = 85, label.y = 4, size = 5) +
  labs(x='Total richness of consumers', y='Consumer aggregated stability', title='Consumer (zooplankton)')

```

```{r benthic macroinvertebrates, fig.width=10, fig.height=4}
## I should remove taxon_code column
## only three lakes but each lake more sites
##
data_binvert_trout <- read.csv('harmonized_data/benthic_macroinvertebrates_trout_lake_area.csv')
# one lake has multiple locations; need to deal with that
## data_binvert_trout$taxon_code <- NULL
##
## remove the rows with species name = empty: 11 rows
data_binvert_trout <- data_binvert_trout[data_binvert_trout$taxon_name!="", ]
#
sum(duplicated(data_binvert_trout[,1:12])==TRUE)
## 0 duplicated rows
##
## convert it into a data.frame: this is a yearly dataset
data_binvert_trout_convert <- data_binvert_trout %>% pivot_wider(names_from=taxon_name, values_from = abundance, values_fill = 0)
## change NA values into 0
data_binvert_trout_convert[is.na(data_binvert_trout_convert)] <- 0.0
##
## aggregated into annual average: no need 
data_binvert_trout_annual <- data_binvert_trout_convert %>% ungroup %>% dplyr::select(-c(site, taxa_type, ecosystem, habitat_fine, biome, guild, plot, unit_abundance, scale_abundance)) 
##
## calculate stability
data_binvert_trout_annual$total <- rowSums(data_binvert_trout_annual[, -(1:5)])
stability_cv <- data_binvert_trout_annual %>% group_by(unique_ID) %>% summarise(cv_1 = mean(total)/sd(total))
##
## further step to site level data
data_binvert_trout_site <- data_binvert_trout_annual %>% ungroup %>% dplyr::select(-year) %>% group_by(unique_ID) %>% summarise_all(mean)
H <- diversity(data_binvert_trout_site[,-c(1:4)])
S <- specnumber(data_binvert_trout_site[,-c(1:4)])
## they are highly correlated. 
plot(H, S)
##
diversity <- data.frame(unique_ID = data_binvert_trout_site$unique_ID, shannon = H, richness = S, lake_id=substring(data_binvert_trout_site$unique_ID, 1, 2))
##
## combine diversity and stability
diversity_stability_binvert <- diversity %>% left_join(stability_cv, by="unique_ID")

plot(data=diversity_stability_binvert, shannon ~ cv_1)
plot(data=diversity_stability_binvert, richness ~ cv_1)
##
## adding the effect of ecosystem size
lake_feature <- read.csv('lake_characteristics.csv')
diversity_stability_binvert <- diversity_stability_binvert %>% left_join(lake_feature, by='lake_id')
plot(data=diversity_stability_binvert, log(surface.area..hectare.) ~ cv_1)    # stability is highly correlated to lake area
##
diversity_stability_binvert$area_log <- log(diversity_stability_binvert$surface.area..hectare.)
summary(lm(data=diversity_stability_binvert, cv_1 ~ area_log + shannon))
summary(lm(data=diversity_stability_binvert, cv_1 ~ area_log))           
summary(lm(data=diversity_stability_binvert, cv_1 ~ shannon))           # this is the best model, but the relationship is negative. 
##
p1 <- ggplot(data=diversity_stability_binvert, aes(x=shannon, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Consumer shannon diversity', y='Consumer aggregated stability (1/cv)', title='Consumer (Benthic invertebrates)')

p2 <- ggplot(data=diversity_stability_binvert, aes(x=richness, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Consumer richness', y='Consumer aggregated stability (1/cv)', title='Consumer (Benthic invertebrates)')

p3 <- ggplot(data=diversity_stability_binvert, aes(x=area_log, y=cv_1)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Log transformed lake surface area', y='Consumer aggregated stability (1/cv)', title='Consumer (Benthic invertebrates)')
##
ggarrange(p1, p2, p3, nrow=1, ncol=3)
ggsave("project_results/consumer_benthic_invertebrate_diversity_stability.png")


```


```{r cross trophic level relationship, fig.width=10, fig.height=8}
diversity_stability_across <- data.frame(unique_ID = diversity_stability_phyto$unique_ID, lake_id = diversity_stability_phyto$lake_id, 
                                         area_log = diversity_stability_phyto$area_log, shannon_phyto = diversity_stability_phyto$shannon, 
                                         richness_phyto = diversity_stability_phyto$richness, cv_1_phyto = diversity_stability_phyto$cv_1)
#
diversity_stability_across <- diversity_stability_across %>% left_join(diversity_stability_zoop[,c("lake_id", "richness", "shannon", "cv_1")], by='lake_id') %>% rename("richness_zoop"="richness", "shannon_zoop"="shannon", "cv_1_zoop"="cv_1")
#
p1 <- ggplot(data=diversity_stability_across, aes(x=shannon_phyto, y=cv_1_zoop)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Producer shannon diversity', y='Consumer aggregated stability (1/cv)', title='Producer-Consumer (zooplankton)')
p1
#
p2 <- ggplot(data=diversity_stability_across, aes(x=shannon_zoop, y=cv_1_phyto)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Consumer shannon diversity', y='Producer aggregated stability (1/cv)', title='Producer-Consumer (zooplankton)')
p2
#
p3 <- ggplot(data=diversity_stability_across, aes(x=shannon_phyto, y=shannon_zoop)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Producer shannon diversity', y='Consumer shannon diversity', title='Producer-Consumer (zooplankton) diversity')
p3
#
p4 <- ggplot(data=diversity_stability_across, aes(x=cv_1_phyto, y=cv_1_zoop)) +
  geom_point() +
  geom_smooth(method='lm') +
  stat_cor(color='green') +
  labs(x='Producer aggregated stability (1/cv)', y='Consumer aggregated stability (1/cv)', title='Producer-Consumer (zooplankton) stability')
p4
##
ggarrange(p1, p2, p3, p4, nrow=2, ncol=2)
ggsave("project_results/producer_consumer_diversity_stability.png")

```


